---
title: "Exercises third lecture"
author: "Giulia Nicole Baldrighi"
date: "2024-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Data import

**NB** remember the libraries
```{r}
library(readxl)
dati <- read_excel("dati_chirurgia.xlsx")
```

### Check for variable names
```{r}
colnames(dati)
```



### Glimpse
```{r,  warning = FALSE}
library(dplyr)
glimpse(dati)
```


### Check for distribution and missing data
```{r}
library(ggplot2)
ggplot(dati, aes(x =as.factor(Comorbidità), fill=Sex)) +
  geom_bar() +
  labs(title = "Barplot Example", x = "Groups", y = "Values") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

**Question**: *does it have sense to include a variable with so many missing values?* 


### NAs
```{r}
library(kableExtra)
nas <- dati %>%
  summarise_all(~ sum(is.na(.)))


nas %>%
  kbl() %>%
  kable_styling()
```

- - - 
### Frequencies of categories
```{r,  warning = FALSE}
library(tidyverse)

lapply(dati, typeof)

table<- table(dati$Comorbidità)

table%>%
  kbl() %>%
  kable_styling()

```

**Question**: *does it have sense to include an under-representative category?*


### Try actively on the data (Comorbidità and Complicanze)


- - - 

### What important information is missing?
```{r}
dati$ID <- paste0("id", 1:110)
dati$ID
```


- - - 


### Unsupervised exploratory analysis 
```{r}
library("FactoMineR")
library("factoextra")

res<- FAMD (dati[,c(3,4,6,7,8,22)], ncp = 3, sup.var =9, ind.sup = NULL, graph = TRUE)


fviz_screeplot(res)
```


```{r}
fviz_mca_var(res)
```



```{r}
# Contribution to the first dimension
fviz_contrib(res, "var", axes = 1)
```


```{r}
# Contribution to the second dimension
fviz_contrib(res, "var", axes = 2)
```


```{r}
fviz_contrib(res, "var", axes = 3)
```


```{r}

fviz_famd_var(res, "quali.var", col.var = "contrib", 
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
)

```






