---
title: "Exercises second lecture"
author: "Giulia Nicole Baldrighi"
e-mail: "giulianicole.baldrighi01@universitadipavia.it"
date: "2024-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  ## edited options below
  fig.width = 8,
  fig.asp = 0.8,
  out.width = "100%",
  dpi = 300
)
```


## Data 
We will use a collection of data for practicing in medicine (`medicaldata`). 

```{r}
library(medicaldata)
```

Today's lecture has a practical session on data from the following study on cytomegalovirus infection. 
In this dataset data from patients who underwent T-cell replete is collected, with respect to matched sibling donor reduced-intensity conditioning allogeneic hematopoietic stem
cell transplant. 

In this study, the primary risk factor of interest was the number of activating killer immunoglobulin-like receptors (aKIRs: 1-4 vs. 5-6). A number of demographic,
baseline and transplant characteristics were also collected. The primary outcome is
presence of and time to cytomegalovirus reactivation. The dataset is cleaned and
relatively complete. There are no outliers or data problems. These are data from a
study by Sobecks et al. “Cytomegalovirus Reactivation After Matched Sibling Donor
Reduced-Intensity Conditioning Allogeneic Hematopoietic Stem Cell Transplant
Correlates With Donor Killer Immunoglobulin-like Receptor Genotype”. Exp Clin Transplant 2011; 1: 7-13.




## First of all, a bit of statistics

**What does it mean that we do not have outliers?**

- Usually outliers may compromise inference if we are interesetd in some model of prediction. 

- Good practice to perform that is to check for normality assumption of data, via tests of normality, namely, the Kolmogorov–Smirnov test and the Shapiro–Wilk test.


Let's assume we have normal data and non-normal data.

```{r}
normal_data <- rnorm(300)

#create data that follows an exponential distribution
non_normal_data <- rexp(300, rate=3)

```


```{r}
#create histogram for both datasets
par(mfrow=c(1,2))
hist(normal_data, col='steelblue', main='Normal')
hist(non_normal_data, col='steelblue', main='Non-normal')
```
**Shapito test**

```{r}
shapiro.test(normal_data)
```

```{r}
shapiro.test(non_normal_data)
```
**QQplot**

```{r}
qqnorm(normal_data, main='Normal')
qqline(normal_data)
```


```{r}
qqnorm(non_normal_data, main='Normal')
qqline(non_normal_data)
```


**Kolmogorov-Smirnov**
```{r}
ks.test(normal_data, 'pnorm')
```
```{r}
ks.test(non_normal_data, 'pnorm')
```




## Practicing on the data

We start by importing the dataset cytomegalovirus, by creating an internal object in R.

```{r}
dati2 <- medicaldata::cytomegalovirus
```


### Steps: reading the data

### 1) How many and which variables are there?


```{r}
length(colnames(dati2))
```


```{r}
colnames(dati2)
```
```{r}
typeof(dati2$sex)
dati2$aKIRs<- as.factor(dati2$aKIRs)
```


### 2) Cross sectional or longitudinal data? 

Which are the differences?


**Cross-sectional data**

- one time point

- only different individuals in the sample (no repetitions)

- snapshot of a given point in time 


**Longitudinal data**

- several points in time

- same individuals sampled more than one time 

- time varies



### 3) Are there missing data?

```{r}
sapply(dati2,
       function(x) sum(is.na(x)))
colSums(is.na(dati2))
```

### 4) Create a data frame with the information on missing data per variable: 

**Way one**: 
```{r}
table.missing<- as.data.frame(sapply(dati2,
       function(x) sum(is.na(x))))

colnames(table.missing)<- "NAs"
table.missing
```

**Way two**: 
```{r}
table.missing2<- data.frame(colSums(is.na(dati2)))
colnames(table.missing2)<- "NAs"
table.missing2
```


### kableExtra
```{r, warning=FALSE}

library(kableExtra)

table.missing %>%
  kbl() %>%
  kable_paper("hover", full_width = F)
```



```{r}
table2 <- dati2[1:10,2:7]

table2 %>%
  kbl(caption = "Recreating booktabs style table") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```



```{r}
table2 %>%
  kbl() %>%
  kable_minimal()
```



```{r, warning=FALSE, messages=FALSE}
library(ggplot2)
library(dplyr)
```

## Plots and exploratory statistics

**Note**
Human leucocyte antigen (HLA) typing on donors and recipients was performed to
allow assessment of killer immunoglobulin-like receptor ligands (KIRs). To allow for
comparison with previous studies, donors were categorized as having 1-4 or 5-6
activating killer immunoglobulin-like receptor genes (aKIRs). 


### Histogram
```{r}
# Libraries
library(tidyverse)
library(hrbrthemes)



# plot

  ggplot(data=dati2, aes(x=time.to.cmv)) +
    geom_histogram( binwidth=3, 
                    fill="#69b3a2", color="#e9ecef", alpha=0.9) +
    ggtitle("Bin size = 3") +
    theme_ipsum() +
    theme(
      plot.title = element_text(size=15)
    )

```



```{r}
dati2.a <- dati2 %>% 
  filter(aKIRs < 5)
  
ggplot(data = dati2, mapping = aes(x = time.to.cmv )) +
  geom_histogram(binwidth = 2, fill="red", col="black")+
  ggtitle("Time to cytomegalovirus")
```

### Calculate the mean, median and standard deviation of the time to cytomegalovirus.

```{r}
all<- data.frame(mean=mean(dati2$time.to.cmv),
                        median=median(dati2$time.to.cmv),
                        sd=sd(dati2$time.to.cmv))
```



### Now let's try to answer to these questions: 

**1)** Are there differences between the *time.to.cytomegalovisus*
 in patients with 1<=aKIRs<=4 and those ones with aKIRs >4?
 
 - Explore with graphic distribution and calculate means, medians and standard deviations. 
 
 - which kind of test could you aplly to verify this hypothesis?

 
 ### Graphical distributions:
 
Patients with aKIRs 1-4: 
```{r}
dati2.a <- dati2 %>% 
  filter(aKIRs < 5)
  
ggplot(data = dati2.a, mapping = aes(x = time.to.cmv))+
  geom_histogram(binwidth = 2, fill="lightblue", col="black")+
  ggtitle("Time to cytomegalovirus in aKIRs < 5")
```
 
 
Patients with aKIRs 5-6:
```{r}
dati2.b <- dati2 %>% 
  filter(aKIRs > 4)
  
ggplot(data = dati2.b, mapping = aes(x = time.to.cmv)) +
  geom_histogram(binwidth = 2, fill="orange",  col="black")+
  ggtitle("Time to cytomegalovirus in aKIRs >4")
```

 
Patients with aKIRs 1-4:
```{r}
group.a<- data.frame(mean=mean(dati2.a$time.to.cmv),
                        median=median(dati2.a$time.to.cmv),
                        sd=sd(dati2.a$time.to.cmv))
  
  

```
 
Patients with aKIRs 5-6:
```{r}
group.b<- data.frame(mean=mean(dati2.b$time.to.cmv),
                        median=median(dati2.b$time.to.cmv),
                        sd=sd(dati2.b$time.to.cmv))
  
```


```{r}
dati2$akirs.group<- 0
dati2$akirs.group[dati2$aKIRs==5 | dati2$aKIRs==6] <- "b"
dati2$akirs.group[dati2$aKIRs!=5 & dati2$aKIRs!=6] <- "a"

```



### Boxplot

```{r}
ggplot(data = dati2, mapping = aes(x = as.factor(akirs.group), y = time.to.cmv, fill=as.factor(akirs.group))) +
  geom_boxplot(fill=c('lightblue', 'darkred'))+
   guides(fill = guide_legend(title = "Sex"))+
  xlab("aKIRs group")
```

**Are they equal (considered statistically similar)?**
```{r}
wilcox.test(dati2.a$time.to.cmv, dati2.b$time.to.cmv)
```


Divided per sex:
```{r}
ggplot(data = dati2, mapping = aes(x = as.factor(sex), y = time.to.cmv, fill=as.factor(sex))) +
   geom_boxplot(fill=c('lightblue', 'darkred'))+
   guides(fill = guide_legend(title = "Sex"))+
   xlab("Sex")
```

Dividing per sex in group a: 
```{r}

dati2.am <- dati2.a[dati2.a$sex==0, ]
  
dati2.af <- dati2.a[dati2.a$sex==1, ]


# wilcox.test.default(dati2.am$time.to.cmv, dati2.af$time.to.cmv)
```
Why is result different than before?


Dividing per sex in group b: 
```{r}

dati2.bm <- dati2.b[dati2.b$sex==0, ]
  
dati2.bf <- dati2.b[dati2.b$sex==1, ]


wilcox.test(dati2.bm$time.to.cmv, dati2.bf$time.to.cmv)
```
Why is p value even greater than before?


### Barplots


Patients with akirs 5-6:

```{r}
ggplot(data = dati2) +
  geom_bar(mapping = aes(x = prior.chemo), fill="blue")+
  theme_light()
```



### Scatter plots

To visualize trends (linear, polinomial)

```{r}
ggplot(data = dati2, mapping = aes(x = aKIRs, y = time.to.cmv) )+
  geom_point()+
  theme_bw()
```


```{r}
ggplot(data = dati2, mapping = aes(x = aKIRs, y = time.to.cmv, colour=as.factor(sex))) + 
  geom_point()+
  guides(fill = guide_legend(title = "Sex"))
```

### Other types of boxplots
Distribution of quantitative variables
```{r}

ggplot(data = dati2, mapping = aes(x = as.factor(aKIRs), y = time.to.cmv, fill=as.factor(sex))) +
  geom_boxplot()+
   guides(fill = guide_legend(title = "Sex"))

```



```{r}
bp <- ggplot(dati2, aes(x=as.factor(sex), y=time.to.cmv, group=as.factor(sex))) + 
  geom_boxplot(aes(fill=as.factor(sex)))+
   guides(fill = guide_legend(title = "Sex"))

bp

bp + facet_grid(as.factor(donor.cmv) ~ .)

# Split in horizontal direction
bp + facet_grid(. ~ as.factor(donor.cmv))
```

```{r}
bp + facet_grid(as.factor(aKIRs) ~ .)
# Split in horizontal direction
bp + facet_grid(. ~ as.factor(aKIRs))
```

### dataReporter

```{r}
library(dataReporter)
```



```{r}
#makeDataReport(dati2, )
```



