---
title: "Exercises first lecture"
author: "Giulia Nicole Baldrighi"
date: "2024-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing packages if not present

```{r, echo=FALSE}
#install.packages("tidyverse")
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("hrbrthemes")
```


### Hands on Tidyverse R syntax

```{r, warning=FALSE, message=FALSE}
```


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```


**Data import**

```{r}
data <- read.csv("C:/Users/gnbal/Desktop/GitHub/RVerse/data/dati.csv.gz")
```


**1) How many different mice were profiled in the experiment?**

Number of samples (mice):

```{r}
samples<- unique(data$Sample_ID)
length(samples)
```


**2) How many observations per mouse? Was there an equal number of observations for each mouse?**

Observations per mouse:
```{r}
data %>% count(Sample_ID)
```


**3) Is there any missing data? How much? Is there a mouse with more missing data than other mice? Or a week, a day, a ZT time with more missing data?**

There are different ways:
```{r}
#base R
summary(data) 
sapply(data,
       function(x) sum(is.na(x)))
colSums(is.na(data))
```


Missing per week:
```{r}
# tidyr
data %>%
    #group_by(Sample_ID) %>%
        group_by(Week) %>% 
        #group_by(Day) %>%
        #group_by(ZT) %>%
    summarise_all(~sum(is.na(.)))

```


Missign per day:
```{r}
# tidyr
data %>%
    #group_by(Sample_ID) %>%
        #group_by(Week) %>% 
        group_by(Day) %>%
        #group_by(ZT) %>%
    summarise_all(~sum(is.na(.)))
```


Missing per ZT:
```{r}
# tidyr
data %>%
    #group_by(Sample_ID) %>%
        #group_by(Week) %>% 
        #group_by(Day) %>%
        group_by(ZT) %>%
    summarise_all(~sum(is.na(.)))
```


- - -

**4) Across how many weeks was the experiment performed?**

Weeks of observation:
```{r}
data %>% count(Week)
```


**5) What is the mean, SD, and range in glucose value across weeks and days between the two different groups of mice?**

Glucose through different groups:
```{r, warning=FALSE, message=FALSE}
data %>%
       group_by(Group,Week,Day) %>%
       summarise(mean = mean(Glucose),
                 sd = sd(Glucose),
                 range = range(Glucose))

data %>%
    group_by(Group,Week,Day) %>%
    summarise(mean = mean(Glucose,na.rm=TRUE),
              sd = sd(Glucose,na.rm=TRUE),
              range = paste(range(Glucose),collapse=" - "))

data %>%
    group_by(Group,Week,Day) %>%
    summarise(across(Glucose,
                     .f = list(mean = mean,
                               sd = sd,
                               min = min,
                               max = max),
                     na.rm = TRUE))
```



