---
title: "Hands on data"
author: "Giulia Nicole Baldrighi"
date: "2024-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Library
```{r}
library(medicaldata)
```

Dataset
```{r}
data<- medicaldata::polyps
```

**Dataset description** 

22 participants were assigned to 1 of 2 treatments, sulindac or placebo.

The initial sample consisted of 22 people with known FAP.

N = 22 subjects (2 dropped out between month 3 and 12) 7 variables

- participant_id

- sex

- age

- baseline (number of polyps at baseline visit)

- treatment (sulindac or placebo)

- number3m (number of polyps at 3 months)

- number6m (number of polyps at 6 months)


**Pivot wider format**
```{r}
head(data)
```



### Question: Is there a difference between placebo and treated groups?

```{r}
colnames(data)
```


```{r}
colnames(data)[4] <- "time0"
colnames(data)[6:7] <- c("time1", "time2")
```


**Pivot longer format**
```{r, warning=FALSE}
library(tidyr)
long_data <- pivot_longer(data, 
                           cols = starts_with("time"),  # Specify columns to pivot
                           names_to = "time",  # Name of new variable storing column names
                           values_to = "number")    # Name of new variable storing values

```


```{r}
head(long_data)
```


### Plots
```{r, warning=FALSE}
library(ggplot2)


ggplot(long_data, aes(x = time, y = number, fill = factor(time))) +
  geom_bar(stat = "identity") +
  facet_wrap(~ participant_id, scales = "free_x") +
  labs(x = "Time", y = "Number of polyps ", fill= "Time") +
  ggtitle("Patients")+
  theme_minimal()
```

```{r}
ggplot(long_data, aes(x = participant_id, y = number, fill = factor(time))) +
  geom_bar(stat = "identity") +
  labs(x = "Patient", y = "Number of polyps", fill="Time") +
  theme_minimal()
```

```{r}
ggplot(long_data, aes(x = participant_id, y = number, fill = factor(time))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Patient", y = "Number of polyps", fill="Time") +
  theme_minimal()

```

```{r, warning=FALSE}
library(dplyr)
library(car)


```

### Estimate of treatment

**GEE**
```{r, warning=FALSE}
library(gee)
library(MuMIn)
```
```{r}
#Modelli GEE
fm <- gee(number ~   time + treatment, id = participant_id,
           data =long_data, corstr = "unstructured")

```


```{r}
summary(fm)
```



Significance
```{r}
#p-value
Robustz<- c(2.78, -1.30, -1.26)
1-pnorm(abs(Robustz))
```



**ANOVA with repeated measurments**
```{r}
library(ez)
```



```{r}
model <- aov(number ~ time+ treatment + Error(participant_id/time), data = long_data)
summary(model)
```
**Mean difference**
```{r}
mean_diff <- aggregate(number ~ time + treatment, data = long_data, FUN = mean)
```


```{r}
summary(mean_diff)
```


